Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2023-07-21 src/docs_statml-poster/statml-poster.Rmd
Poster update! Nicer figure 2, get rid of big table, better PCA-AGHQ description. Need just to add some new results in now #57

* District-level model of HIV indicators which synthesises data from 1) household surveys, 2) antenatal care (ANC) clinics, and 3) routine service provision of antiretroviral therapy (ART) [@eaton2021naomi]
  * Combining evidence from multiple data sources helps overcome the limitations of any one
  * Small-area estimation methods (structured random effects) to overcome limited district-level sample sizes
* Yearly estimation process: model run interactively by by country teams using `naomi.unaids.org` web-app
  * Figure \@ref(fig:fig1) illustrates the seven stages of using the app
* Inference conducted in minutes using empirical Bayes and a Gaussian approximation
* Days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no]: not practical!
* Naomi has a large latent field $x$ controlled by a smaller number of hyperparameters $\theta$
* Extended latent Gaussian model [@stringer2022fast]: more complex dependency structures than a latent Gaussian model [@rue2009approximate]
* Looking for a fast, approximate approach, that properly takes uncertainty in hyperparameters into account

@@

* **Laplace approximation** Integrate out latent field using a Gaussian approximation to the denominator
$$
p(\theta, y) \approx \tilde p_\texttt{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\texttt{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)},
$$
where $\tilde p_\texttt{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \hat H(\theta)^{-1})$
  * Use automatic differentiation via `CppAD` in `TMB`

```{r aghq, echo=FALSE, fig.align='center', fig.cap='Demonstration of PCA-AGHQ.'}
knitr::include_graphics("fig2.png")
```

* **Adaptive Gauss-Hermite Quadrature** (AGHQ) perform quadrature over the hyperparameters
$$
\int_\Theta p_\texttt{LA}(\theta, y) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p_\texttt{LA}(\hat \theta + Lz, y) \omega(z),
$$
where the Gauss-Hermite quadrature rule $\{z \in \mathcal{Q}(m, k), \omega\}$ with $m = \dim(\theta)$ and $k$ points per dimension is adapted based upon
  * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p_\texttt{LA}(\theta, y)$
  * A matrix decomposition $LL^\top = - \partial_\theta^2 \log p_\texttt{LA}(\theta, y) \rvert_{\theta = \hat \theta}$
* Use the spectral decomposition $L = E\Lambda^{1/2}$ and keep only the first $s < m$ **principal components** (PCA-AGHQ)

@@

| Method | Description | Time |
|-----------|----------|------|
| TMB | Baseline | `r round(time_taken$TMB, 0)` secs |
| PCA-AGHQ | Ours | `r round(time_taken$aghq, 0)` hour |
| NUTS | Gold-standard | `r round(time_taken$tmbstan, 1)` days |

@@

ks_reduction <- ks %>%
  filter(type == "Latent field") %>%
  summarise(
    ks_tmb = mean(`KS(TMB, tmbstan)`),
    ks_aghq = mean(`KS(aghq, tmbstan)`)
  ) %>%
  mutate(reduction = 100 * (ks_aghq - ks_tmb) / ks_tmb) %>%
  pull(reduction) %>%
  round()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# second90 <- readr::read_csv("depends/second90.csv")
# 
# second90_reduction <- second90 %>%
#   summarise(
#     rmse_tmb = 100 * sqrt(mean((tmbstan - TMB)^2)),
#     rmse_aghq = 100 * sqrt(mean((tmbstan - aghq)^2))
#   ) %>%
#   mutate(reduction = 100 * (rmse_aghq - rmse_tmb) / rmse_tmb) %>%
#   pull(reduction) %>%
#   round()
```

* For PCA-AGHQ $k = 3$ and $s = 8$ chosen using Scree plot to explain ~90% of variance
* For NUTS 4 chains of 100,000 thinned by 40 were required for good diagnostics
* Kolmogorov-Smirnov (KS) test based on the maximum difference between marginal ECDFs
  * Average KS distance from NUTS reduced by `r abs(ks_reduction)`%
  * Also considering joint posteriors via Pareto-smoothed importance sampling and maximum mean discrepancy
* Naomi can be used to assess probabilities targets have been met e.g. 90% of those who know their HIV status are on ART ("second 90"). Both TMB and PCA-AGHQ have biased inferences

<!-- ```{r fig4, echo=FALSE, fig.align='center', fig.cap='Both approximate methods are meaningfully incorrect for policy.'} -->
<!-- knitr::include_graphics("fig4.png") -->
<!-- ``` -->

>>>>

* District-level model of HIV indicators which synthesises data from 1) household surveys, 2) antenatal care (ANC) clinics, and 3) routine service provision of antiretroviral therapy (ART) [@eaton2021naomi]
  * Combining evidence from multiple data sources helps overcome the limitations of any one
  * Small-area estimation methods (structured random effects) to overcome limited district-level sample sizes
* Yearly estimation process: model run interactively by by country teams using `naomi.unaids.org` web-app
  * Figure \@ref(fig:fig1) illustrates the seven stages of using the app
* Inference conducted in minutes using empirical Bayes and a Gaussian approximation
* Days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no]: not practical!
* Naomi has a large latent field $x$ controlled by a smaller number of hyperparameters $\theta$
* Extended latent Gaussian model [@stringer2022fast]: more complex dependency structures than a latent Gaussian model
* Looking for a fast, approximate approach, that properly takes uncertainty in hyperparameters into account

@@

```{r aghq, echo=FALSE, fig.align='center', fig.cap='Demonstration of Gauss-Hermite quadrature, adaption, and our principal components approach.'}
knitr::include_graphics("fig2.png")
```

* **Laplace approximation** Integrate out latent field using a Gaussian approximation to the denominator
$$
p(\theta, y) \approx \tilde p_\texttt{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\texttt{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)},
$$
where $\tilde p_\texttt{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \hat H(\theta)^{-1})$
  * Use automatic differentiation via `CppAD` in `TMB`
* **Adaptive Gauss-Hermite Quadrature** (AGHQ) perform quadrature over the hyperparameters
$$
\int_\Theta p_\texttt{LA}(\theta, y) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p_\texttt{LA}(\hat \theta + Lz, y) \omega(z),
$$
where $m = \dim(\theta)$ and the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ with weights $\omega:\mathcal{Q} \mapsto \mathbb{R}$ and $k$ points per dimension is adapted based upon
  * The mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p_\texttt{LA}(\theta, y)$
  * A matrix decomposition $LL^\top = - \partial_\theta^2 \log p_\texttt{LA}(\theta, y) \rvert_{\theta = \hat \theta}$
* **PCA-AGHQ** To integrate over large spaces, use the spectral decomposition $L = E\Lambda^{1/2}$ and keep only the first $s < m$ principal components

@@

@@

ks_reduction <- ks %>%
  filter(type == "Latent") %>%
  summarise(
    ks_tmb = mean(TMB),
    ks_aghq = mean(aghq)
  ) %>%
  mutate(reduction = 100 * (ks_aghq - ks_tmb) / ks_tmb) %>%
  pull(reduction) %>%
  round()
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# second90 <- readr::read_csv("depends/second90.csv")
# 
# second90_reduction <- second90 %>%
#   summarise(
#     rmse_tmb = 100 * sqrt(mean((tmbstan - TMB)^2)),
#     rmse_aghq = 100 * sqrt(mean((tmbstan - aghq)^2))
#   ) %>%
#   mutate(reduction = 100 * (rmse_aghq - rmse_tmb) / rmse_tmb) %>%
#   pull(reduction) %>%
#   round()
```
* Run three methods: 1. TMB (baseline, `r round(time_taken$TMB, 0)` secs), 2. PCA-AGHQ (new, `r round(time_taken$aghq, 0)` hour), 3. NUTS (gold-standard, `r round(time_taken$tmbstan, 1)` days)
  * For PCA-AGHQ $k = 3$ and $s = 8$ chosen using Scree plot to explain ~90% of variance
  * For NUTS 4 chains of 100,000 thinned by 40 were required for good diagnostics e.g. all $\hat R < 1.025$
* Naomi can be used to assess probabilities targets have been met e.g. 90% of those who know their HIV status are on ART ("second 90"). Both TMB and PCA-AGHQ have biased inferences
